---
title: "Precipitation Event Clustering: Robustness Analysis for different Minimum Interevent Times"
subtitle: "Minimum Interevent Time (IET) = `r params$iet` hours"
author: "Lena Ortega Menjivar"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    fig_width: 10
    fig_height: 6
params:
  iet: 6
  stations: !r c(30, 171)
  path_results: "data/clusres/"
  path_dat: "data/events_combined/"
  filt: 0
  ks: !r 2:8
  clusvars: "magnitude"
  regvars: "magnitude"
  chosen_ks: !r NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center',
  dpi = 300
)

library(tidyverse)
library(ggpubr)
library(flexmix)
library(FactoMineR)
library(philentropy)
library(knitr)
devtools::load_all()

theme_set(theme_minimal(base_size = 12))
```

```{r load-data, include=FALSE}
# Parameters
iet <- params$iet
stations <- params$stations
pths <- function(path) {
  ifelse(startsWith(path, '/'), #in the offline case, where the source files are stored separately
         path, paste0('../../', path)) #in the 'online' repo/HPC case, relative to .Rmd
}
path_dat <- pths(params$path_dat)
path_results <- pths(params$path_results)
filt <- params$filt
ks <- params$ks
clusvars <- params$clusvars
regvars <- params$regvars
chosen_ks <- params$chosen_ks

# Load data
dat <- sprintf("%smodel_input_events_iet_%dh_%d.RDS",
        path_dat, iet, stations) |> 
  sapply(readRDS, simplify=FALSE) %>%
  setNames(str_extract(names(.), "\\d+(?=\\.RDS)"))|> 
  lapply(\(df) {
  mutate(df,
         flash = as.logical(flash),
         across(all_of(clusvars[1:4]), log),
         across(all_of(clusvars[1:5]), ~(. - min(.))/diff(range(.)))) |> 
    rename(precipitation=rr)
    })

# Load clustering results
part <- new.env() #already contains stability analysis
mod <- new.env()
modStab <- new.env()

ld2env <- function(pathvec, env, pattern="\\d+(?=\\.RDA)") {
  for(i in pathvec) {
    stn <- str_extract(i, pattern = pattern)
    before <- ls(env)
    load(i, envir = env)
    obj_name <- setdiff(ls(env), before)   # find what was just loaded
    assign(stn, get(obj_name, envir = env), envir = env)
    rm(list = obj_name, envir = env)
  }
}

res <- sprintf("%spart_iet_%d_station_%d.RDA",
               path_results, iet, stations)
ld2env(res, part)
parts <- ls(envir=part)

res <- sprintf("%smod_iet_%d_filt_%g_station_%d.RDA",
               path_results, iet, filt, stations)
ld2env(res, mod)
mods <- ls(envir=mod)

res <- sprintf("%smod_iet_%d_filt_%g_station_%d_stabAn.RDA",
               path_results, iet, filt, stations)
ld2env(res, modStab, "\\d+(?=\\_stabAn.RDA)")
modStabs <- ls(envir=modStab)
```

# Executive Summary

This report presents the results of precipitation event clustering and clusterwise regression analysis, and then consecutive cluster-aggregate theoretical distribution fitting for **IET = `r iet` hours**.

**Analysis Configuration:**

- **Inter-Event Time (IET):** `r iet` hours
- **Stations Analyzed:** `r paste(stations, collapse = ", ")`
- **Number of Clusters (k) tested:** `r paste(ks, collapse = ", ")`
- **Filter Values:** `r paste(filt, collapse= ", ")`
- **Date Generated:** `r format(Sys.Date(), '%B %d, %Y')`

---

# 1. Clustering Stability Analysis

## 1.1 Adjusted Rand Index (ARI) - Global Stability of clustering solutions

```{r ari-global-part, fig.height=4, fig.width=7, echo=FALSE}
lapply(parts, \(mdl) {
  partres <- part[[mdl]]
  partres$globalStability |> 
    ggplot(aes(x=k, y=ARI)) +
    geom_boxplot() +
    labs(title=sprintf("station: %s", mdl))
}) |> ggarrange(plotlist=_) |> 
  annotate_figure(top="Partitioning clustering")
```
```{r ari-global-mod, fig.height=4, fig.width=7, echo=FALSE}
lapply(modStabs, \(mdl) {
  regres <- modStab[[mdl]]
  regres$rand |> 
    as.data.frame() |> 
    pivot_longer(everything(),
                 names_to = "k",
                 values_to = "ARI") |> 
    ggplot(aes(x=k, y=ARI)) +
    geom_boxplot() +
    labs(title=sprintf("station: %s", mdl))
}) |> ggarrange(plotlist=_) |> 
  annotate_figure(top="Clusterwise regression")
rm(list=ls(pattern="modStab"))
```

## 1.2 Adjusted Rand Index (ARI) - Segment-level Stability of clustering solutions

Segment-level stability analysis examines the stability of each segment. Only run for partitioning clustering. 

```{r ari-segment, fig.height=8, fig.width=7, echo=FALSE}
lapply(parts, \(mdl) {
  partres <- part[[mdl]]
  partres$segLevelStability |> 
    ggplot(aes(x=segment, y=ARI)) +
    geom_boxplot() +
    facet_wrap(~k, scales="free_x") +
    labs(title=sprintf("station: %s", mdl))
}) |> ggarrange(plotlist=_) |> 
  annotate_figure(top="Partitioning clustering")

kmssg <- ifelse(is.null(chosen_ks),
                "**CAUTION, For the automatized reports the ks chosen for IET=4h are chosen as a starting default, and will possibly have to be adjusted manually.**",
                "**CAUTION, The optimal k was chosen visually based on the above ARI plots, rerun with different values if preferred.**")

```

`r kmssg`

---

# 2. Visualization of clusters

## 2.1 FAMD plots

```{r famd-plots, fig.height=8, fig.width=7, echo=FALSE}

if(is.null(chosen_ks)) {
  chosen_ks <- c(3, 3, 4, 2) |> as.character()
  names(chosen_ks) <- rep(c("part", "mod"), each=2) |> 
    paste0(stations)
}

part <- sapply(parts, \(mdl) {
  partres <- part[[mdl]]$stepClustResults
  stn <- str_extract(mdl, "\\d+$")
  clus_short <- .get_clusters(partres,
                              chosen_ks[paste0("part", stn)])
  tms <- dat[[stn]]$precipitation |> 
    sapply(\(x) sum(x>filt))
  clus_long <- rep(clus_short, times=tms)
  
  list(clus_short=clus_short,
       clus_long=clus_long)
  
}, simplify=FALSE)

mod <- sapply(mods, \(mdl) {

  regres <- mod[[mdl]]
  stn <- str_extract(mdl, "\\d+$")
  kstar <- chosen_ks[paste0("mod", stn)]
  mod <- getModel(regres, kstar)
  clus_short <- .get_clusters(regres, kstar)
  clus_long <- clusters(mod) #same as .get_clusters(regres, per_group=FALSE)
  
  list(mod=mod,
       clus_short=clus_short,
       clus_long=clus_long)
  
}, simplify=FALSE)

allmod <- data.frame(
  mod_type=c(rep('part', length(part)),
             rep('mod', length(mod))),
  station=c(names(part), names(mod))
) |> 
  mutate(mdl=paste0(mod_type, station)) |> 
  left_join(enframe(chosen_ks, name='mdl', value='chosen_k'))
clusses <- setNames(c(part, mod),
                    allmod$mdl)

FAMD_DIMS <- list(x=1, y=2:4)

apply(allmod, 1, \(i) {
  stn <- i['station']
  mdl <- i['mdl']
  mod_type <- i['mod_type']
  kstar <- i['chosen_k']
  plt_famd(dat[[stn]][clusvars],
           clusses[[mdl]]$clus_short,
           dims=FAMD_DIMS,
           title=sprintf("Station: %s; Mod.Type: %s; k=%s", stn, mod_type, kstar))
})  %>% ggarrange(plotlist=.,
                 legend.grob=get_legend(.[[which.max(chosen_ks)]] +
                                          theme(legend.position = 'bottom')),
                 legend='bottom')

```


# 2.2. Partial Dependence Plots (PDP)

Marginal effect of each (main effect of a) predictor on the response variable, applied only to clusterwise regression results.

```{r pdp-plots, fig.height=6, fig.width=7, echo=FALSE}

lapply(mods, \(mdl) {
  mod_type <- "mod"
  kstar <- filter(allmod, mod_type==mod_type &
                    station==mods[1])$chosen_k
  plt_dpd_flexmix(mod[[mdl]]$mod,
                  setdiff(regvars, "flash"),
                  fac_var="flash",
                  title=sprintf("Station: %s; Mod.Type: %s; k=%s", mdl, mod_type, kstar))
}) %>% ggarrange(plotlist=., ncol=1,
                 legend.grob=get_legend(.[[which.max(chosen_ks[grep("mod", names(chosen_ks))])]] +
                                          theme(legend.position = 'bottom')),
                 legend='bottom')

```


## 2.3 Clusterwise Empirical Cumulative Distribution Functions (ECDF)

```{r ecdf-plots, fig.height=6, fig.width=7, echo=FALSE}

lapply(stations, \(stn) {
  i <- allmod$station==stn
  mdls <- allmod$mdl[i]
  lbls <- paste("Type:", allmod$mod_type[i]) |> 
    setNames(mdls)
  dat[[as.character(stn)]] |> 
    rename(rr=precipitation) |>
    plt_ecdf(clus_obj = lapply(clusses[mdls], `[[`, "clus_short"),
           title=paste("Station:", stn),
           subtitle=lbls)
}) |> ggarrange(plotlist=_, ncol=1,
               common.legend=TRUE,
               legend='bottom')

```

---

# 3. Cluster-Aggregate distribution fit and comparison

# 2.1 Kullback-Leibler Divergence Analysis


```{r kl-divergence, results='asis', echo=FALSE}

LENGTH_OUT <- 1000

KL <- apply(allmod, 1, \(i) {
  
  stn <- i['station']
  mdl <- i['mdl']
  rr <- dat[[stn]]$precipitation
  
  pdf_theoretical <- 
    aggregate_dist_calc(rr, clusses[[mdl]]$clus_long,
                        filt=filt,
                        length_out = LENGTH_OUT,
                        type='PDF')
  rr <- unlist(rr)
  rr <- rr[rr>filt]
  pdf_empirical <- density(rr,
                           from = max(filt, 1e-3),
                           to = max(rr),
                           n = LENGTH_OUT)$y
  
  c(
    simple_Gamma=.get_KL(pdf_empirical, pdf_theoretical$simple_Gamma),
    aggregated_Gamma=.get_KL(pdf_empirical, pdf_theoretical$aggregated_Gamma)
  )
})

KL |> t() |> 
  kable()

```


# 3.2 Visual CDF Comparison

These plots compare the fits of the simple theoretical distribution and the clusterwise-aggregate theoretical distribution to the empirical CDF of the precipitation time series.


```{r cdf-comparison, fig.height=6, fig.width=7, echo=FALSE}

lapply(stations, \(stn) {
  i <- allmod$station==stn
  mdls <- allmod$mdl[i]
  lbls <- paste("Type:", allmod$mod_type[i]) |> 
    setNames(mdls)
  dat[[as.character(stn)]] |> 
    rename(rr=precipitation) |>
    plt_cdf_comp(clus_vec = lapply(clusses[mdls], `[[`, "clus_short"),
                 filt=filt,
                 title=paste("Station:", stn),
                 subtitle=lbls)
  }) |> ggarrange(plotlist=_, ncol=1,
               common.legend=TRUE,
               legend='bottom')

```

# Session Information

```{r session-info}
sessionInfo()
```

---

**Report generated:** `r Sys.time()`  
**IET value:** `r iet` hours  
**Stations:** `r paste(stations, collapse = ", ")`
